---
title: "1. Data Cleaning"
subtitle: "MY498: Capstone Project 2025-26"
author: "68508"
date: today
format:
    html:
        embed-resources: true
        toc: true
        number-sections: true
        fontsize: 12pt
---
# Setup
This notebook constructs the analysis dataset used in all subsequent notebooks. It downloads raw data programmatically from official sources, cleans and reshapes each dataset, and assembles a balanced borough-level quarterly panel covering 2011 Q4 to 2024 Q4.
All raw data downloaded to `data/raw` and processed outputs are saved to `data/processed/`. Neither folder is tracked by Git - the notebook is fully self-contained and reproducible from scratch.

## Packages 
Downloading and ensuring all required packages exist 
```{r}
#| label: packages
#| message: false
#| warning: false

required_packages <- c(
    # Data wrangling
    "tidyverse", # dplyr, tidyr, ggplot2, readr, purrr, stringr
    "readODS", # read .ods files (DVLA tables)
    "readxl", # read .xls/.xlsx files
    "janitor", # clean_names(), tabyl()
    "lubridate", # date handling for quarterly periods
    "zoo", # yeargtr class for quarterly time series
    # Spatial / GIS
    "sf", # shape files, spatial joins, distance calc
    "units", # unit-aware distances (metres/km)
    # Econometrics 
    "fixest", #feols() - fast fixed effects with clustered SEs (Callaway-Sant'Anna)
    "did", # Callaway & Sant'Anna (2021) staggered DiD estimator
    "rdrobust", # RDD: MSE-optimal bandwidth, robust CIs
    "rddensity", # McCrary density test at the boundary
    "tidysynth", # synthetic control
    # Output / tables
    "modelsummary", # regression tables
    "kableExtra", # formatted HTML/LaTeX tables
    "ggthemes", # clean plot themes
    # Utilities 
    "curl", # internet connectivity check 
    "here", # project-relative paths
    "nomisr" # programmatic access to ONS/OMIS data
)

# Install missing packages
installed <- rownames(installed.packages())
to_install <- required_packages[!required_packages %in% installed]

if (length(to_install) > 0) {
    install.packages(to_install, repos = "https://cloud.r-project.org")
}

if (!"nomisr" %in% rownames(installed.packages())) {
    if(!"remotes" %in% rownames(installed.packaages())) {
        install.packages("remotes", repos = "https://cloud.r-project.org")
    }
    remotes::install_github("ropensci/nomisr")
}

# Load all packages
invisible(lapply(required_packages, library, character.only = TRUE))

message("All packages loaded successfully")
```

## Paths & Download Helper

```{r}
#| label: paths

data <- here("data")
raw <- here("data", "raw")
Proc <- here('data', 'processed')

# Create folders if it doesn't exist
if (!dir.exists(data)) {
    dir.create(data)
}

if (!dir.exists(raw)){
    dir.create(raw)
}
if (!dir.exists(Proc)) {
    dir.create(Proc)
}

# Helper function to download if files are not present
download_if_missing <- function(url, dest) {
    if (!file.exists(dest)) {
        message("Downloading ", basename(dest), "...")
        download.file(url, dest, mode='wb', quiet = TRUE)
        message("Saved: ", dest)
    } else {
        message("Already present: ", basename(dest))
    }
}
```

# Data Construction
## Spatial Data

### London Borough Boundaries
We load the London borough boundary shapefile from the London Datastore. The Shapefile uses the British National Grid project (EPSG: 27700) which gives distances in metres - essential for the running variable construction. We compute borough centroids as the representative point for each borough, used to measure distance to the ULEZ boundaries.

```{r}
#| label: borough-boundaries
#| message: false

borough_spatial <- "https://data.london.gov.uk/download/20od9/9ba8c833-6370-4b11-abdc-314aa020d5e0/statistical-gis-boundaries-london.zip"

download_if_missing(borough_spatial, file.path(raw, "london_boundaries.zip"))

# Unzip files 
boundaries_dir <- file.path(raw, "london_boundaries")
if (!dir.exists(boundaries_dir)){
    unzip(file.path(raw, "london_boundaries.zip"), exdir = boundaries_dir)
    message("Unzipped boundaries.")
}

boroughs_sf <- st_read(
    file.path(boundaries_dir, "statistical-gis-boundaries-london/ESRI/London_Borough_Excluding_MHW.shp")
) |>
    clean_names() |>
    rename(ons_code = gss_code, la_name=name) |>
    select(ons_code, la_name, geometry) |>
    st_set_crs(27700) |>
    st_make_valid()
```

```{r}
# Compute centroids
boroughs_sf <- boroughs_sf |>
    mutate(
        centroid = st_centroid(geometry),
        centroid_x = st_coordinates(centroid)[,1],
        centroid_y = st_coordinates(centroid)[,2]
    )

centroids_sf <- boroughs_sf |>
    st_drop_geometry() |>
    st_as_sf(coords = c("centroid_x", "centroid_y"), crs = 27700)

message("Boroughs loaded: ", nrow(boroughs_sf))
```

### Surrounding Districts - Never-Treated Control Units
To strengthen identification for the 2023 expansion analysis, we include local authority districts that directly border Greater London as additional never-treated control units. These districts share the same commuter spillover exposure as outer London boroughs - their residents face ULEZ charges when driving into London - but were never subject to the policy expansion themselves.
We identify bordering districts programmatically using spatial intersection with the Greater London boundary.

```{r}
#| label: surrounding-districts
#| message: false

# Greater London boundary - union of all boroughs
london_boundary <- boroughs_sf |> st_union()
eng_path <- file.path(raw, "england_districts.geojson")

# Download full England district boundaries from ONS ArcGIS API 
# File is ~63MB - cached locally after first successful download
# If download fails,the chunk falls back to a hardcoded list of bordering districts identified via spatial analysis on 2026-07-05
if (!file.exists(eng_path)) {
    message("Attempting to download England district boundaries...")
    tryCatch({
        library(httr2)
        resp <- request(paste0(
            "https://services1.arcgis.com/ESMARspQHYMw9BZ9/arcgis/rest/services/",
            "Local_Authority_Districts_May_2024_Boundaries_UK_BFC/FeatureServer/0/query",
            "?where=LAD24CD+LIKE+'E%25'",
            "&outFields=LAD24CD,LAD24NM",
            "&returnGeometry=true",
            "&f=geojson"
        )) |>
            req_timeout(600) |>
            req_retry(max_tries = 3, backoff = ~30) |>
            req_perform()
        writeBin(resp_body_raw(resp), eng_path)
        message("Downloaded and cached: england_districts.geojson")
    }, error = function(e) {
        message("Download failed: ", conditionMessage(e))
        message("Falling back to hardcoded district list.")
    })
}

# Path 1. Load from cached full dataset and identify programmatically 
if (file.exists(eng_path)) {
    england_districts <- st_read(eng_path, quiet = TRUE)
    surrounding_sf <- england_districts |>
        filter(!LAD24CD %in% boroughs_sf$ons_code) |>
        filter(lengths(st_is_within_distance(
            geometry, london_boundary, dist = 2000)) > 0 ) |>
        distinct(LAD24CD, .keep_all = TRUE) |>
        rename(ons_code = LAD24CD, la_name = LAD24NM)
    message("Surrounding districts identified programmatically: ", nrow(surrounding_sf))

# Path 2: Fallback -targ
} else {
    message("Using targeted query for known bordering districts...")
    surrounding_codes <- c(
        "E07000068", # Brentwood
        "E07000095", # Broxbourne
        "E06000060",  # Buckinghamshire
        "E07000107",  # Dartford
        "E07000207",  # Elmbridge
        "E07000072",  # Epping Forest
        "E07000208",  # Epsom and Ewell
        "E07000098",  # Hertsmere
        "E07000210",  # Mole Valley
        "E07000211",  # Reigate and Banstead
        "E07000111",  # Sevenoaks
        "E06000039",  # Slough
        "E07000213",  # Spelthorne
        "E07000215",  # Tandridge
        "E07000102",  # Three Rivers
        "E06000034",  # Thurrock
        "E07000103",  # Watford
        "E07000241",  # Welwyn Hatfield
        "E06000040"   # Windsor and Maidenhead
    )
    surrounding_query <- paste0(
        "https://services1.arcgis.com/ESMARspQHYMw9BZ9/arcgis/rest/services/",
        "Local_Authority_Districts_May_2024_Boundaries_UK_BFC/FeatureServer/0/query",
        "?where=LAD24CD+IN+(",
        paste0("'", surrounding_codes, "'", collapse=","),
        ")&outFields=LAD24CD,LAD24NM&returnGeometry=true&f=geojson"
    )
    download_if_missing(surrounding_query, file.path(raw, "surrounding_districts.geojson"))
    surrounding_sf <- st_read(file.path(raw, "surrounding_districts.geojson"), quiet=TRUE) |>
        st_transform(27700) |>
        rename(ons_code = LAD24CD, la_name = LAD24NM)
}

saveRDS(surrounding_sf, file.path(Proc, "surrounding_sf.rds"))
message("Saved: surrounding_sf.rds")

study_codes <- c(as.character(boroughs_sf$ons_code), as.character(surrounding_sf$ons_code))
message("Total study units: ", length(study_codes),
" (", length(boroughs_sf$ons_code), " boroughs + ", length(surrounding_sf$ons_code), " surrounding districts)")
```

### ULEZ Boundaries 
We download the three ULEZ expansion boundaries and compute signed distance running variables for all 52 study units. The signed distance is negative for units inside the zone and positive for units outside.

```{r}
#| label: ulez-boundaries-setup
#| message: false

study_sf <- bind_rows(
    boroughs_sf |> select(ons_code, la_name, geometry),
    surrounding_sf |> select(ons_code, la_name, geometry)
)

study_sf <- study_sf |>
    mutate(
        centroid = st_centroid(geometry),
        centroid_x = st_coordinates(centroid)[,1],
        centroid_y = st_coordinates(centroid)[,2]
    )

centroids_sf <- study_sf |>
    st_drop_geometry() |>
    st_as_sf(coords=c("centroid_x", "centroid_y"), crs = 27700)

message("Study units with centroids: ", nrow(study_sf))
```
```{r}
#| label: ulez-2019
#| message: false

boundary_2019 <- "https://data.london.gov.uk/download/v8onw/84a78e36-7c27-4ba0-868f-13a76e1b50cb/ULEZCentral_CongestionChargingZone.zip"

download_if_missing(boundary_2019, file.path(raw, "ulez_2019.zip"))

ulez_2019_dir <- file.path(raw, "ulez_2019")
if (!dir.exists(ulez_2019_dir)){
    unzip(file.path(raw, "ulez_2019.zip"), exdir = ulez_2019_dir)
    message("Unzipped ulez_2019.")
}

ulez_2019 <- st_read(
    file.path(ulez_2019_dir, "Shapefile/UltraLowEmissionsZoneBoundary(ULEZ).shp")
) |>
    st_set_crs(27700) |>
    st_make_valid()

# Convert the ULEZ 2019 polygon to a line so we can measure distance 
# to the boundary edge rather than to the interior of the polygon
ulez_2019_line <- st_cast(st_union(ulez_2019), "MULTILINESTRING")

dist_2019 <- st_distance(centroids_sf, ulez_2019_line)

study_sf <- study_sf |>
    mutate(
        inside_2019 = as.integer(lengths(st_intersects(study_sf, ulez_2019)) > 0
        ),
        dist_to_2019_km = as.numeric(dist_2019[,1]) / 1000,
        rv_2019 = ifelse(inside_2019==1,
        -dist_to_2019_km, dist_to_2019_km)
    )

message("Boroughs inside 2019 zone: ", sum(study_sf$inside_2019))
```

```{r}
#| label: ulez-2021
#| message: false

boundary_2021 <- "https://data.london.gov.uk/download/vdjx4/56a082c1-4626-4d6f-80fe-d7fa99d8fdd5/InnerUltraLowEmissionZone.zip"

download_if_missing(boundary_2021, file.path(raw, "ulez_2021.zip"))

ulez_2021_dir <- file.path(raw, "ulez_2021")
if (!dir.exists(ulez_2021_dir)) {
    unzip(file.path(raw, "ulez_2021.zip"), exdir = ulez_2021_dir)
    message("Unzipped ulez_2021.")
}

ulez_2021 <- st_read(file.path(ulez_2021_dir, "InnerUltraLowEmissionZone.shp")) |>
    st_set_crs(27700) |>
    st_make_valid()
```

```{r}
# Convert to line for distance calculation
ulez_2021_line <- st_cast(st_union(ulez_2021), "MULTILINESTRING")

# Distance from each borough centroid to the 2021 boundary line
dist_2021 <- st_distance(centroids_sf, ulez_2021_line)

# Add to study_sf
study_sf <- study_sf |>
    mutate(
        inside_2021 = as.integer(lengths(st_within(centroids_sf, ulez_2021)) > 0
        ),
        dist_to_2021_km = as.numeric(dist_2021[,1]) / 1000,
        rv_2021 = ifelse(inside_2021 == 1,
        -dist_to_2021_km, dist_to_2021_km)
    )

message("Boroughs inside 2021 zone: ", sum(study_sf$inside_2021))
```

```{r}
#| label: ulez-2023
#| message: false

download_if_missing("https://data.london.gov.uk/download/vd455/2e43ce0a-5ae3-41cb-be60-e33b662f5470/londonWideUltraLowEmissionZone.zip", file.path(raw, "ulez_2023.zip")
)

ulez_2023_dir <- file.path(raw, "ulez_2023")
if (!dir.exists(ulez_2023_dir)) {
    unzip(file.path(raw,"ulez_2023.zip"), exdir = ulez_2023_dir)
    message("Unzipped ulez_2023.")
}
```

```{r}
ulez_2023 <- st_read(file.path(ulez_2023_dir, "londonWideUltraLowEmissionZone.shp")) |>
    st_set_crs(27700) |>
    st_make_valid()

ulez_2023_line <- st_cast(st_union(ulez_2023), "MULTILINESTRING")

dist_2023 <- st_distance(study_sf, ulez_2023_line)

study_sf <- study_sf |>
    mutate(
        dist_to_2023_km = as.numeric(dist_2023[,1]) / 1000,
        rv_2023 = ifelse(
            lengths(st_within(centroids_sf, ulez_2023)) > 0,
            -dist_to_2023_km, dist_to_2023_km
        ),
        treated = as.integer(rv_2021 > 0)
    )

message("Treated boroughs (2023 expansion): ", sum(study_sf$treated))
message("Control borough (inside 2021 zone): ", sum(study_sf$treated==0))
```

```{r}
saveRDS(study_sf, file.path(Proc, "study_sf.rds"))
message("Saved: study_sf.rds — ", nrow(study_sf), " units")
```


```{r}
#| label: clean-spatial
#| message: false

rm(list = setdiff(ls(), c(
    # Data
    "study_sf",
    "surrounding_sf",
    "study_codes",
    
    "ulez_2019", "ulez_2021", "ulez_2023",
    # Paths
    "raw", "Proc",
    
    # Helper function
    "download_if_missing",
    
    # Package management
    "required_packages", "installed", "to_install"
)))

gc()
message("Objects retained: ", paste(ls(), collapse = ", "))
```

## ULEV Compliance Share

### VEH0132 - Licensed ULEVs by Local Authority, Quarterly

VEH0132 is the primary outcome variable. It records the number of licensed ultra low emission vehicles by local authority, fuel type, and keepership on a quarterly basis. We filter to all 52 study units (33 London boroughs + 19 surrounding districts), pivot from wide to long format, and aggregate across fuel types and keepership to produce total ULEV counts per unit per quarter.

We retain disaggregated counts by fuel types (BEV vs PHEV) and keepership (company vs private) to enable heterogeneity analysis in subsequent notebooks. BEV adoption is a stronger signal of anticipatory compliance than PHEV since it requires full commitment to electric-only driving. Company fleet compliance may differ from private given that the fleet replacement cycles are planned in advance.

```{r}
#| label: veh0132
#| message: false

 veh0132 <- "https://assets.publishing.service.gov.uk/media/69ef35519ca985145673b9d7/veh0132.ods"

 download_if_missing(veh0132, file.path(raw, "veh0132.ods"))

 raw_ulev <- read_ods(
    file.path(raw, "veh0132.ods"),
    sheet = "VEH0132",
    skip=4
 )
```

```{r}
raw_ulev <- clean_names(raw_ulev)

# Identify quarter columns - named "x2025_q4" etc after clean_names()
qtr_cols <- grep("^x\\d{4}_q\\d$", names(raw_ulev), value=TRUE)
message("Quarter columns found: ", length(qtr_cols), " | ", tail(qtr_cols, 1), " to ", qtr_cols[1])

raw_ulev <- raw_ulev |>
    mutate(across(all_of(qtr_cols), as.character))

# Filter to London boroughs, pivot to long and clean counts 
ulev_long <- raw_ulev |>
    filter(ons_code %in% study_codes) |>
    filter(fuel != "Total", keepership != "Total") |>
    select(-units, -ons_sort) |>
    pivot_longer(
        cols = all_of(qtr_cols),
        names_to = "quarter_raw",
        values_to = "ulev_count"
    ) |>
    mutate(
        year = as.integer(str_extract(quarter_raw, "\\d{4}")),
        quarter = as.integer(str_extract(quarter_raw, "\\d$")),
        yq = as.yearqtr(paste0(year, " Q", quarter)),
        ulev_count = case_when(
            ulev_count %in% c("[z]", "[x]", "[low]", "") ~ NA_character_,
            TRUE ~ ulev_count),
        ulev_count = as.numeric(str_remove_all(ulev_count, ","))
        ) |>
    select(ons_code, ons_geography, fuel, keepership, year, quarter, yq, ulev_count)
```

```{r}
ulev_total <- ulev_long |>
    group_by(ons_code, ons_geography, year, quarter, yq) |>
    summarise(
        ulev_total   = sum(ulev_count,                                    na.rm = TRUE),
        ulev_bev     = sum(ulev_count[fuel == "BATTERY ELECTRIC"],        na.rm = TRUE),
        ulev_phev    = sum(ulev_count[str_detect(fuel, "PLUG-IN")],       na.rm = TRUE),
        ulev_company = sum(ulev_count[keepership == "COMPANY"],           na.rm = TRUE),
        ulev_private = sum(ulev_count[keepership == "PRIVATE"],           na.rm = TRUE),
        .groups      = "drop"
    )

glimpse(ulev_total)
```

```{r}
saveRDS(ulev_total, file.path(Proc, "ulev_total.rds"))
message("Saved: ulev_total.rds — ", nrow(ulev_total), " rows | ",
        n_distinct(ulev_total$ons_code), " study units")
```

### VEH0105: All Licensed Vehicle by Local Authority, Quarterly 

VEH0105 provides the denominator for our primary outcome variable. It records the total number of licensed vehicles by local authority, body type, fuel type, and keepership on a quarterly basis. We filter to cars only and to all 52 study units, then aggregate to unit-quarter total car counts. Note that VEH0105 reports counts in thousands — we multiply by 1,000 to convert to actual vehicle counts for consistency with VEH0132.

```{r}
#| label: veh0105 - clean
#| message: false

veh0105 <- "https://assets.publishing.service.gov.uk/media/69ef3553ed93f72cf81633fc/veh0105.ods"

download_if_missing(veh0105, file.path(raw, "veh0105.ods"))

raw_veh <- read_ods(
    file.path(raw, "veh0105.ods"),
    sheet = "VEH0105",
    skip = 4
)
```

```{r}
qtr_cols_raw_veh <- grep("^\\d{4} Q\\d$", names(raw_veh), value = TRUE)
message("Quarter columns found: ", length(qtr_cols_raw_veh), " | ", tail(qtr_cols_raw_veh, 1), " to ", qtr_cols_raw_veh[1])

raw_veh <- clean_names(raw_veh)

qtr_cols_veh <- grep("^x\\d{4}_q\\d$", names(raw_veh), value = TRUE)

raw_veh <- raw_veh |>
    mutate(across(all_of(qtr_cols_veh), as.character))
```
```{r}
#| label: veh0105-clean
#| message: false

# Filter to london boroughs
cars_long <- raw_veh |>
    filter(ons_code %in% study_codes,
        body_type == "Cars",
        fuel != "Total",
        keepership != "Total") |>
    select(-units, -ons_sort) |>
    pivot_longer(
        cols = all_of(qtr_cols_veh),
        names_to = "quarter_raw", 
        values_to = "count_raw"
    ) |>
    mutate(
        year = as.integer(str_extract(quarter_raw, "\\d{4}")),
        quarter = as.integer(str_extract(quarter_raw, "\\d$")),
        yq = as.yearqtr(paste0(year, " Q", quarter)),
        count_raw = case_when(
            count_raw %in% c("[z]", "[x]", "[low]", "") ~ NA_character_,
            TRUE ~ count_raw
        ),
        cars = as.numeric(str_remove_all(count_raw, ",")) * 1000
    ) |>
    select(ons_code, ons_geography, fuel, keepership, year, quarter, yq, cars)

# Aggregate to borough x quarter total cars 
cars_total <- cars_long |>
    group_by(ons_code, ons_geography, year, quarter, yq) |>
    summarise(
        cars_total = sum(cars, na.rm=TRUE),
        cars_company = sum(cars[keepership == "COMPANY"], na.rm = TRUE),
        cars_private = sum(cars[keepership == "PRIVATE"], na.rm = TRUE),
        .groups = "drop"
    )
glimpse(cars_total)
```

```{r}
saveRDS(cars_total, file.path(Proc, "cars_total.rds"))
message("Saved: cars_total.rds — ", nrow(cars_total), " rows | ",
        n_distinct(cars_total$ons_code), " study units")
```

### Compliance share 
The compliance share is the primary outcome variable - the proportion of licensed cars in each study unit that are ultra low emission vehicles. We merge `ulev_total` and `cars_total` on borough and quarter, compute the compliance share and disaggregated shares by fuel type and keepership, and add time indicators for the announcement and implementation dates.

```{r}
#| label: compliance-share
#| message: false

panel_compliance <- ulev_total |>
    left_join(cars_total,
        by = c("ons_code", "ons_geography", "year", "quarter", "yq")) |>
    mutate(
        ulev_share = ulev_total/cars_total,
        bev_share = ulev_bev / cars_total,
        phev_share = ulev_phev / cars_total,
        company_share = ulev_company / cars_company,
        private_share = ulev_private / cars_private,
        post_announce = as.integer(yq >= as.yearqtr("2022 Q4")),
        post_implement = as.integer(yq >= as.yearqtr("2023 Q3")),
        anticipation_window = as.integer(yq >= as.yearqtr("2022 Q4") & yq < as.yearqtr("2023 Q3")),
        t_announce = (year - 2022) * 4 + (quarter - 4)
    )

glimpse(panel_compliance)
```

```{r}
saveRDS(panel_compliance, file.path(Proc, "panel_compliance.rds"))

message("Saved: panel_compliance.rds — ", nrow(panel_compliance), " rows | ",
        n_distinct(panel_compliance$ons_code), " study units")
```

```{r}
# =========================================
# Clean Environment 
# =========================================
rm(list = setdiff(ls(), c(
    # Data 
    "panel_compliance",
    "study_sf",
    "surrounding_sf",
    "study_codes",
    "ulev_total",
    "cars_total",
    "ulez_2019",
    "ulez_2021",
    "ulez_2023",
    # Paths
    "raw",
    "Proc",
    # Helper function 
    "download_if_missing",
    # Package management
    "required_packages",
    "installed",
    "to_install"
)))

gc()

message("Objects retained: ", paste(ls(), collapse = ", "))
```


## Control Variables
### Car/Van Availability (TS045)

Car ownership is a key cross-sectional control - study units with higher baseline car ownership are more likely to have residents who need to comply with ULEZ. We download Census 2021 TS045 from the Nomis API, which provides household counts by number of cars or vans available.

```{r}
#| label: car-availability
#| message: false

# T045 - car/van availability by local authority
# NM_2063_1, categories 1-4 (excluding total), counts
car_avail_raw <- nomis_get_data(
    id = "NM_2063_1", 
    time = "latest",
    geography = "TYPE424",
    C2021_CARS_5 = "1,2,3,4",
    measures = 20100,
    select = c("GEOGRAPHY_NAME", "GEOGRAPHY_CODE", "C2021_CARS_5_NAME", "OBS_VALUE")
)
```

```{r}
car_avail <- car_avail_raw |>
    clean_names() |>
    filter(geography_code %in% study_codes) |>
    pivot_wider(
        names_from = c2021_cars_5_name,
        values_from = obs_value
    ) |>
    clean_names() |>
    rename(
        ons_code = geography_code, 
        la_name = geography_name
    ) |>
    mutate(
        total_households = no_cars_or_vans_in_household + x1_car_or_van_in_household + x2_cars_or_vans_in_household + x3_or_more_cars_or_vans_in_household,

        pct_no_car = no_cars_or_vans_in_household / total_households, 
        pct_car_owner = 1 - pct_no_car, 

        avg_cars_hh = (0 * no_cars_or_vans_in_household + 1* x1_car_or_van_in_household + 2 * x2_cars_or_vans_in_household + 3.5 * x3_or_more_cars_or_vans_in_household) / total_households
    ) |>
    select(ons_code, la_name, total_households, pct_no_car, pct_car_owner, avg_cars_hh)

glimpse(car_avail)
```

```{r}
saveRDS(car_avail, file.path(Proc, "car_avail.rds"))
message("Saved: car_avail.rds — ", nrow(car_avail), " rows")
```

### Household Deprivation (TS011)
Deprivation is a cross-sectional control for socioeconomic status. We download Census 2021 TS011 from the Nomis API, which provides household counts by number of deprivation dimensions (0-4).

```{r}
#| label: deprivation
#| message: false

# TS011 - Household deprivation by local authority 
# NM_2031_1, categories 1-5 (excluding total), counts
deprivation_raw <- nomis_get_data(
    id = "NM_2031_1",
    time = "latest",
    geography = "TYPE424",
    C2021_DEP_6 = "1,2,3,4,5",
    measures = 20100,
    select = c("GEOGRAPHY_NAME", "GEOGRAPHY_CODE", "C2021_DEP_6_NAME", "OBS_VALUE")
)

deprivation <- deprivation_raw |>
    clean_names() |>
    filter(geography_code %in% study_codes) |>
    pivot_wider(
        names_from  = c2021_dep_6_name,
        values_from = obs_value
    ) |>
    clean_names() |>
    rename(
        ons_code = geography_code,
        la_name  = geography_name
    ) |>
    mutate(
        # Total households from parts
        total_hh = household_is_not_deprived_in_any_dimension +
                   household_is_deprived_in_one_dimension +
                   household_is_deprived_in_two_dimensions +
                   household_is_deprived_in_three_dimensions +
                   household_is_deprived_in_four_dimensions,
        # Share deprived in at least one dimension
        pct_deprived = 1 - (household_is_not_deprived_in_any_dimension / 
                             total_hh),
        # Weighted deprivation index — more dimensions = higher weight
        deprivation_index = (1 * household_is_deprived_in_one_dimension +
                             2 * household_is_deprived_in_two_dimensions +
                             3 * household_is_deprived_in_three_dimensions +
                             4 * household_is_deprived_in_four_dimensions) /
                             total_hh
    ) |>
    select(ons_code, la_name, pct_deprived, deprivation_index)

glimpse(deprivation)
```

```{r}
saveRDS(deprivation, file.path(Proc, "deprivation.rds"))
message("Saved: deprivation.rds — ", nrow(deprivation), " rows")
```

```{r}
saveRDS(deprivation, file.path(Proc, "deprivation.rds"))
message("Saved: deprivation.rds")
```

### Population Density (TS006)
Population density is a cross-sectional control for urban density. We download Census 2021 TS006 from the Nomis API.

```{r}
#| label: pop-density
#| message: false

pop_density_raw <- nomis_get_data(
    id        = "NM_2026_1",
    time      = "latest",
    geography = "TYPE154",
    measures  = 20100
)
pop_density <- pop_density_raw |>
    clean_names() |>
    filter(geography_code %in% study_codes) |>
    rename(
        ons_code = geography_code,
        la_name = geography_name,
        pop_density_per_km2 = obs_value
    ) |>
    select(ons_code, la_name, pop_density_per_km2)
glimpse(pop_density)
```

```{r}
saveRDS(pop_density, file.path(Proc, "pop_density.rds"))
message("Saved: pop_density.rds — ", nrow(pop_density), " rows")
```

### Median Annual Earnings (ASHE)

Median annual earnings is a time-varying control for income levels. We download the Annual Survey of Hours and Earnings (ASHE) resident analysis from the Nomis API for 2011-2024, covering our full study period.

```{r}
#| label: ashe-earnings
#| message: false

# ASHE — Median annual gross pay by local authority, 2011-2024
# NM_30_1, total sex (7), median (2), annual gross pay (7)
# Downloaded year by year to avoid URL length limits

ashe_raw <- map_dfr(2011:2024, function(yr) {
    nomis_get_data(
        id        = "NM_30_1",
        time      = as.character(yr),
        geography = "TYPE424",
        sex       = 7,
        item      = 2,
        pay       = 7,
        measures  = 20100
    )
})

ashe <- ashe_raw |>
    clean_names() |>
    filter(geography_code %in% study_codes) |>
    rename(
        ons_code = geography_code, 
        la_name = geography_name,
        year = date,
        median_earnings_gbp = obs_value
    ) |>
    select(ons_code, la_name, year, median_earnings_gbp)

glimpse(ashe)
```

```{r}
saveRDS(ashe, file.path(Proc, "ashe_earnings.rds"))
message("Saved: ashe_earnings.rds — ", nrow(ashe), " rows | ",
        n_distinct(ashe$ons_code), " study units")
```

### House Prices (HMLR UK HPI)
Average house prices are a time-varying control for neighbourhood wealth. We download the Land Registry UK House Price Index which provides monthly average prices by local authority back to 1968. We aggregate to quarterly by taking the last month of each quarter.

```{r}
#| label: house-prices
#| message: false

# HMLR UK HPI — Average prices by local authority, monthly

download_if_missing(
    "https://publicdata.landregistry.gov.uk/market-trend-data/house-price-index-data/Average-prices-2026-04.csv",
    file.path(raw, "house_prices.csv")
)

hpi_raw <- read_csv(
    file.path(raw, "house_prices.csv"),
    show_col_types = FALSE
)

hpi <- hpi_raw |>
    clean_names()|>
    filter(area_code %in% study_codes) |>
    mutate(
        year = year(date),
        month = month(date),
        quarter = case_when(
            month %in% c(1,2,3) ~ 1L,
            month %in% c(4,5,6) ~ 2L, 
            month %in% c(7,8,9) ~ 3L,
            month %in% c(10,11,12) ~ 4L
        ),
        yq = as.yearqtr(paste0(year, " Q", quarter))
    ) |>
    group_by(area_code, region_name, year, quarter, yq) |>
    slice_max(date, n=1) |>
    ungroup() |>
    rename(
        ons_code = area_code,
        la_name = region_name,
        house_price = average_price
    ) |>
    select(ons_code, la_name, year, quarter, yq, house_price)

glimpse(hpi)
```

```{r}
saveRDS(hpi, file.path(Proc, "house_prices.rds"))
message("Saved: house_prices.rds — ", nrow(hpi), " rows | ",
        n_distinct(hpi$ons_code), " study units")
```

```{r}
#| label: clean-3
#| message: false

rm(list = setdiff(ls(), c(
    # Data
    "panel_compliance",
    "study_sf",
    "surrounding_sf",
    "study_codes",
    
    # Controls
    "car_avail",
    "deprivation",
    "pop_density",
    "ashe",
    "hpi",
    
    # ULEZ boundaries for plotting
    "ulez_2019",
    "ulez_2021",
    "ulez_2023",
    
    # Paths
    "raw", "Proc",
    
    # Helper function
    "download_if_missing",
    
    # Package management
    "required_packages", "installed", "to_install"
)))

gc()
message("Objects retained: ", paste(ls(), collapse = ", "))
```

## Final Panel assembly 

```{r}
#| label: panel-assembly
#| message: false

study_sf <- study_sf |>
    mutate(
        london = as.integer(str_starts(ons_code, "E09"))
    )

message("London boroughs: ", sum(study_sf$london == 1))
message("Surrounding districts: ", sum(study_sf$london == 0))

# Extract spatial variables from study_sf
study_vars <- study_sf |>
    select(ons_code, treated, london, rv_2019, rv_2021, rv_2023,
           dist_to_2019_km, dist_to_2021_km, dist_to_2023_km,
           centroid_x, centroid_y, geometry)

# Merge everything into one panel

analysis_panel <- panel_compliance |>
    rename(la_name = ons_geography) |>
    left_join(study_vars |> st_drop_geometry(), by = "ons_code") |>
    left_join(car_avail,   by = c("ons_code", "la_name")) |>
    left_join(deprivation, by = c("ons_code", "la_name")) |>
    left_join(pop_density, by = c("ons_code", "la_name")) |>
    left_join(ashe,        by = c("ons_code", "la_name", "year")) |>
    left_join(
        hpi |> select(-la_name),
        by = c("ons_code", "year", "quarter", "yq")
    ) |>
    left_join(study_vars |> select(ons_code, geometry), by = "ons_code") |>
    st_as_sf() |>
    filter(yq <= as.yearqtr("2024 Q4"))

message("Panel rows: ", nrow(analysis_panel))
message("Study units: ", n_distinct(analysis_panel$ons_code))
message("Time range: ", min(analysis_panel$yq), " to ", max(analysis_panel$yq))

saveRDS(analysis_panel, file.path(Proc, "analysis_panel.rds"))
message("Saved: analysis_panel.rds — ", nrow(analysis_panel), " rows, ",
        ncol(analysis_panel), " columns")
```

```{r}
saveRDS(analysis_panel, file.path(Proc, "analysis_panel.rds"))
message("Saved: analysis_panel.rds — ", nrow(analysis_panel), " rows, ",
        ncol(analysis_panel), " columns")
```

## Final Quality Checks 

```{r}
#| label: quality-checks
#| message: false

# 1. Balanced panel check
panel_balance <- analysis_panel |>
    st_drop_geometry() |>
    group_by(ons_code) |>
    summarise(n_quarters = n(), .groups = "drop")

message("Balanced panel check:")
message("  Min quarters per unit: ", min(panel_balance$n_quarters))
message("  Max quarters per unit: ", max(panel_balance$n_quarters))

# 2. Treatment split
message("\nTreatment split:")
message("  London boroughs — treated (outer): ",
        sum(study_sf$treated == 1 & study_sf$london == 1))
message("  London boroughs — control (inner): ",
        sum(study_sf$treated == 0 & study_sf$london == 1))
message("  Surrounding districts (never-treated): ",
        sum(study_sf$london == 0))

# 3. NAs in key variables
message("\nNAs in key variables:")
message("  ulev_share:          ", sum(is.na(analysis_panel$ulev_share)))
message("  house_price:         ", sum(is.na(analysis_panel$house_price)))
message("  median_earnings_gbp: ", sum(is.na(analysis_panel$median_earnings_gbp)))
message("  rv_2021:             ", sum(is.na(analysis_panel$rv_2021)))

# 4. Time range
message("\nTime range: ", min(analysis_panel$yq), " to ", max(analysis_panel$yq))
```